---
title: "Maps with harp"
format: html
editor: visual
---

## Attach the packages we're going to need

```{r packages, message=FALSE}
library(harp)
library(dplyr)
library(tidyr)
library(meteogrid)
```

## Read the data

It's important that the data are in a data frame for ggplot, so `read_forecast()` is the best option. To demonstrate getting the model elevation, surface geopotential ($\Phi_{sfc}$) will be read from the file.

$$
z_{sfc}=\frac{\Phi_{sfc}}{9.80665}
$$

```{r read-data}
sfc_geo <- read_forecast(
  2024013100,
  "meps",
  "sfc_geo",
  lead_time     = 0,
  file_template = "meps_det",
  return_data   = TRUE 
)
```

We can now scale the surface geopotential to height in metres:
```{r}
sfc_geo <- scale_param(sfc_geo, 1 / 9.80665, new_units = "m", mult = TRUE)
```


## Get the map outline

We can get a map outline by getting the domain from the forecast column in the data. By setting `polygon = FALSE`, the map is returned as paths rather than polygons since paths are slightly easier to plot. 

```{r map-outline}
countries <- get_map(get_domain(sfc_geo$fcst), polygon = FALSE)
```

By default, the map outline is from the "world" database in the _maps_ package. Higher resolution maps are available via the _rnaturalearthdata_ and _rnaturalearthhires_ packages. 

```{r hires-map-outline}
countries_hires <- get_map(
  get_domain(sfc_geo$fcst), 
  map     = rnaturalearthhires::countries10,
  polygon = FALSE
)
```


## Plot using ggplot

The `geom_georaster()` geom, that is a harp extension to ggplot can be used to plot the 2d field. 

```{r basic-plot}
ggplot() +
  geom_georaster(aes(geofield = fcst), sfc_geo) +
  geom_path(aes(x, y), countries)
```

By default the axes are stretched to fill the plot window and a margin is added. We can fix this with `coord_equal(expand = FALSE)`, and remove the superfluous axis labels with `theme_harp_map()`. In addition, to speed up plotting the data can be upscaled using the downsample method, whereby pixels are dropped prior to plotting - this is useful for testing, but it may be better not to use it for the final plot. 

```{r sized-plot}
ggplot() +
  geom_georaster(
    aes(geofield = fcst), 
    sfc_geo,
    upscale_factor = 2,
    upscale_method = "downsample"
  ) +
  geom_path(aes(x, y), countries) +
  coord_equal(expand = FALSE) + 
  theme_harp_map()
```

We can also add a better colour scale using `scale_fill_gradientn()`. Note that the limits argument sets the limits of the colour scale - this is just to make the plot look nicer - a specific landmask should probably be used really. The oob argument tells it what to do with data that are outside the limits. 
```{r colour-scale-sized-plot}
ggplot() +
  geom_georaster(
    aes(geofield = fcst), 
    sfc_geo,
    upscale_factor = 2,
    upscale_method = "downsample"
  ) +
  geom_path(aes(x, y), countries) +
  scale_fill_gradientn(
    "Terrain\nHeight", 
    colours = c("skyblue3", terrain.colors(255)), 
    limits = c(1, NA),
    oob = scales::squish
  ) +
  coord_equal(expand = FALSE) + 
  theme_harp_map()
```

## Get model elevation interpolated to stations

Now, let's say we want to add some text to the plot, e.g. for terrain height in Oslo, Bergen, Trondheim and Tromsø. We can get the interpolated terrain height by reading the data and interpolating it. We'll start by filtering the built in station list. 

```{r filter-stations}
my_stations <- filter(
  station_list,
  name %in% c("OSLO-BLINDERN", "BERGEN/FLORIDA", "TRONDHEIM/VOLL", "TROMSO")
)
```

We can now use these stations in the interpolate options for `read_forecast()`
```{r interpolate-to-stations}
stn_hgts <- read_forecast(
  2024013100,
  "meps",
  "sfc_geo",
  lead_time           = 0,
  file_template       = "meps_det",
  transformation      = "interpolate", 
  transformation_opts = interpolate_opts(my_stations),
  return_data         = TRUE 
) |> 
  scale_param(1 / 9.80665, "m", TRUE) |> 
  mutate(parameter = "height")
```

The data are in lat-lon coordinates so we need to project them into the coordinate system of the maps.

```{r project-data}
projection <- get_domain(sfc_geo$fcst)$projection
stn_hgts <- mutate(
  stn_hgts,
  projected = project(lon, lat, proj = projection)
) |> 
  unnest(projected)
```

We can plot these values (rounded so they don't take loads of space) on the map using `geom_text()`
```{r terrain-plot-with-labels}
ggplot() +
  geom_georaster(
    aes(geofield = fcst), 
    sfc_geo,
    upscale_factor = 2,
    upscale_method = "downsample"
  ) +
  geom_path(aes(x, y), countries, colour = "grey40") +
  geom_text(aes(x, y, label = round(fcst)), stn_hgts) + 
  scale_fill_gradientn(
    "Terrain\nHeight", 
    colours = c("skyblue3", terrain.colors(255)), 
    limits = c(1, NA),
    oob = scales::squish
  ) +
  coord_equal(expand = FALSE) + 
  theme_harp_map()
```

## Plotting a subset

The easiest way to plot a subset of the data is to set the limits in coord_equal(). These values are in the projected coordinate system so we need these values. We can use information from the extent of the domain to get the x and y values for each axis.

```{r get-x-and-y}
dom_info <- DomainExtent(get_domain(sfc_geo$fcst))
x <- seq(dom_info$x0, dom_info$x1, dom_info$dx)
y <- seq(dom_info$y0, dom_info$y1, dom_info$dy)
```

We'll also use the higher resolution map outline as we're zoomed in. 

```{r plot-subset}
ggplot() +
  geom_georaster(
    aes(geofield = fcst), 
    sfc_geo,
    upscale_factor = 2,
    upscale_method = "downsample"
  ) +
  geom_path(aes(x, y), countries_hires, colour = "grey40") +  
  geom_text(aes(x, y, label = round(fcst)), stn_hgts) + 
  scale_fill_gradientn(
    "Terrain\nHeight", 
    colours = c("skyblue3", terrain.colors(255)), 
    limits = c(1, NA),
    oob = scales::squish
  ) +
  coord_equal(
    expand = FALSE, xlim = c(x[200], x[500]), ylim = c(y[300], y[600])) + 
  theme_harp_map()
```

## 3-month means
Reading in 3 months of data, with one field per day results in about 740MB for MEPS data. 
```{r 3-month-t2m, message=FALSE}
t2m <- read_forecast(
    seq_dttm(2023090100, 2023113000, "1d"),
    "meps",
    "T2m",
    lead_time     = 12,
    file_template = "meps_det",
    return_data   = TRUE 
) |> 
  scale_param(-273.15, "degC")
```

We can easily get the mean and add it as a column - each row will have the same value. 
```{r mean-t2m}
t2m <- mutate(t2m, fcst_mean = mean(fcst))
```

And we can also add an anomaly from the mean column
```{r anom-t2m}
t2m <- mutate(t2m, fcst_anom = fcst - fcst_mean)
```

We can make a plot as before using ggplot and `geom_georaster()`. Just to demonstrate how it's done we'll use a binned colour scale to plot the mean (note we only want the first row in the data frame).

```{r binned-scale}
ggplot() + 
  geom_georaster(
    aes(geofield = fcst_mean), 
    t2m[1,], 
    upscale_factor = 2, 
    upscale_method = "downsample"
  ) + 
  geom_path(aes(x, y), countries, colour = "grey") + 
  scale_fill_binned(
    bquote(~degree*"C"), 
    low = "white", high = "red", 
    breaks = seq(-12, 20, 2)
  ) +
  coord_equal(expand = FALSE) + 
  theme_harp_map() + 
  theme(legend.key.height = unit(4, "lines"))
```

We can use `geo_points()` to interpolate some of the data to our small list of stations.

```{r t2m-stns}
t2m_stns <- geo_points(t2m, points = my_stations)
```

And we need to reproject
```{r project-t2m-stns}
t2m_stns <- mutate(
  t2m_stns, 
  projected = project(lon, lat, proj = projection)
) |> 
  unnest(projected)
```

And now we can plot the anomaly at our locations on any given day
```{r t2m-with-stns}
ggplot() + 
  geom_georaster(
    aes(geofield = fcst_mean), 
    t2m[1,], 
    upscale_factor = 2, 
    upscale_method = "downsample"
  ) + 
  geom_path(aes(x, y), countries, colour = "grey") +
  geom_text(
    aes(x, y, label = round(fcst_anom)), 
    filter(t2m_stns, fcst_dttm == as_dttm(2023100100))
  ) +
  scale_fill_binned(
    bquote(~degree*"C"), 
    low = "white", high = "red", 
    breaks = seq(-12, 20, 2)
  ) +
  labs(
    title = "Anomaly from SON 2023 mean",
    subtitle = format(
      as_dttm(2023100112), "%H:%M %a %d %b %Y"
    )
  ) +
  coord_equal(expand = FALSE) + 
  theme_harp_map() + 
  theme(legend.key.height = unit(3.5, "lines"))
```

